Variational oblique predictive clustering trees

### 2024-12-13 reports/paper/vspyct.tex
Extra comment
new version

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{pythontex}

\usepackage[acronym]{glossaries}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{bs}{BS}{Brier Score}
\newacronym{pes}{PES}{Public Employment Services}
\newacronym{pex}{PEX}{probability of exit}
\newacronym{spyct}{SPYCT}{oblique predictive clustering tree}
\newacronym{vspyct}{VSPYCT}{variational oblique predictive clustering tree}
\newacronym{sslmtr}{SSL-MTR}{semi-supervised multi-target regression}
\newacronym{pct}{PCT}{predictive clustering tree}
\newacronym{sop}{SOP}{structured output prediction}
\newacronym{fp}{FP}{False Positive}
\newacronym{fn}{FN}{False Negative}
\newacronym{tpr}{TPR}{True Positive Rate}
\newacronym{fpr}{FPR}{False Positive Rate}
\newacronym{auc}{AUC}{Area under the ROC curve}
\newacronym{tp}{TP}{True Positives}
\newacronym{tn}{TN}{True Negatives}
\newacronym{ecdf}{ECDF}{Empirical Cumulative Distribution Function}
\newacronym{isce}{ISCED}{International Standard Classification of Education}

@@

\begin{document}
\begin{frontmatter}
	\title{Variational oblique predictive clustering trees for multi-target regression}
	\author[1,2]{Viktor Andonovikj\corref{cor1}}
	\ead{viktor.andonovikj@ijs.si}
	\author[1]{Sašo Džeroski}
	\ead{saso.dzeroski@ijs.si}
	\author[1,3]{Biljana Mileva Boshkoska}
	\ead{biljana.mileva@ijs.si}
	\author[1]{Pavle Boškoski}
	\ead{pavle.boskoski@ijs.si}
	\address[1]{Jožef Stefan Institute, Jamova cesta 39, 1000 Ljubljana, Slovenia}
	\address[2]{Jožef Stefan International Postgraduate School, Jamova cesta 39, 1000 Ljubljana, Slovenia}
	\address[3]{Faculty of Information Studies in Novo mesto, Ljubljanska cesta 31b, 8000 Novo mesto, Slovenia}
	\cortext[cor1]{Corresponding author.} 
\begin{abstract}

\Glspl{spyct} are semi-supervised multi-target prediction models mainly used for \gls{sop} problems.
They are computationally efficient and when combined in ensembles they achieve state-of-the-art results.
However, one major issue is that it's challenging to interpret an ensemble of \glspl{spyct} without the use of a model-agnostic method.
We propose variational oblique predictive clustering trees, which address this challenge.
The parameters of each split node are treated as distributions and they are learned through the Variational Bayes method. We evaluate the model on several benchmark datasets of different sizes.
The experimental analyses show that a single \gls{vspyct} achieves state-of-the-art or on-par performance with ensemble of standard \glspl{spyct}. We also present a method for extracting feature importance scores from the model.
Finally, we present a method to visually interpret the model's decision making process through analysis of the relative feature importance in each split node.
\end{abstract}
%\maketitle
\end{frontmatter}
\glsresetall

@@

Our proposed \gls{vspyct} framework diverges from traditional ensemble approaches by directly incorporating Bayesian principles into oblique decision trees' architecture.
This model avoids conventional ensemble strategies, which aggregate multiple trees' outputs, for a probabilistic treatment of model parameters, offering refined mechanisms for uncertainty quantification and predictive performance.
The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, modelling these parameters as random variables with Gaussian distributions.
This represents a paradigm shift toward a probabilistic understanding of decision-making processes within tree-based models, significantly enhancing interpretability, especially in uncertainty estimation domains.
The whole framework is given in Figure~\ref{fig:flowchart}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1.0\textwidth]{methodology_flowchart.pdf}
    \caption{Flowchart of the methodology}
    \label{fig:flowchart}
\end{figure}

It sets a new benchmark for interpretable, efficient, and reliable machine learning models.

@@

Our proposed \glspl{vspyct} model addresses these critical challenges by integrating Bayesian principles directly into the decision tree framework.
%Unlike traditional ensemble strategies, \glspl{vspyct} employs a probabilistic treatment of model parameters through the variational Bayes method, which enhances interpretability and provides inherent uncertainty quantification.
This approach maintains the clarity and simplicity of a single tree model while achieving state-of-the-art performance, thus offering a significant improvement over existing methods.
By introducing uncertainty quantification directly into the decision-making process, \glspl{vspyct} enhances the reliability and applicability of the model across various domains where decision certainty is crucial.
%This advancement bridges the gap between theoretical developments and practical applications, marking a significant step forward in making complex decision trees more accessible, interpretable, and effective.

\section{Preliminaries}

\subsection{Variational Bayes method}

@@

In conducting a stochastic analysis, like survival analysis, the process of inferring model parameters hinges on applying Bayes' theorem.
For observations $x$ produced by a system defined by parameters $\theta$, Bayes' theorem is given as:

@@

This integral rarely yields a closed-form solution.
For cases with multiple dimensions, the computational burden often renders Monte Carlo methods infeasible.
The VB approach offers an approximation to this issue by closely mimicking the true posterior $p(\theta|x)$ with an approximate distribution $q_{\omega^*}(\theta)$. This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterized by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
Typically, the mean-field variational family is utilized, assuming independence among the parameters $\omega$. To find the optimal $\omega^*$, one minimises the Kullback–Leibler (KL) divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:

@@

Given the unknown nature of the true posterior, a rearrangement is needed to compute the KL divergence, simplifying to:

\begin{align*}
KL(q_\omega(\theta) \parallel p(\theta|x)) &= \mathbb{E}_q\left[\log \frac{q_\omega (\theta)}{p(\theta|x)}\right] \\
&=\mathbb{E}_q\left[\log q_\omega(\theta)\right]-\mathbb{E}_q\left[\log p(\theta|x)\right]\\
&=\mathbb{E}_q\left[\log q_\omega(\theta)\right]-\mathbb{E}_q\left[\log p(x,\theta)\right]\\
&=\mathbb{E}_q[\log q_\omega(\theta) - \log p(x,\theta)+\log p(x)]\\
&=-\mathbb{E}_q\left[\log q_\omega(\theta) - \log q_\omega(\theta)\right] + \log p(x)\\
&\quad \quad \underbrace{\phantom{\mathbb{E}_q\left[\log q_\omega(\theta) - \log q_\omega(\theta)\right]}}_{\text{ELBO}}
\label{eq:kl_divergence}
\end{align*}

Maximising the ELBO inversely minimises the KL divergence, with $\log p(x)$ remaining constant and excluded from the optimisation.
Typically, the ELBO maximisation criterion is non-convex, with no assurance of global extremum convergence by optimisation algorithms.
Various algorithms have been employed for this challenge, including coordinate ascent variational inference and stochastic variational inference~\cite{hoffman2013stochastic}, with the latter's gradient calculated as per the black-box variational inference~\cite{sashank2018convergence}.
The adoption of an approximate variational distribution introduces bias contingent on the chosen variational family $\mathcal{Q}$, necessitating a decision grounded in empirical evidence or expert knowledge.
Despite this bias, VB's computational efficiency significantly surpasses traditional methods like Monte Carlo integration.
In our work, the ADAM optimiser~\citep{kingma2014adam} is used to facilitate the ELBO optimisation.

\subsection{Semi-supervised Multi-target Regression in Survival Analysis}

Survival analysis primarily focuses on time-to-event data, where not all events are fully observed due to some form of censoring that halts observation before the event transpires.
The most prevalent form of censoring in this context is right-censoring, which occurs when there is incomplete knowledge about the survival time at the end of the follow-up period.
As a result, each survival record is characterised by three elements: the actual survival time $T$, its observed component $t$, and a binary indicator $\delta$, which is defined as:
\begin{equation}
\delta = \begin{cases}
0, & \text{if } T>t \text{ indicating a right-censored record},\\
1, & \text{if } T=t \text{, denoting that the event has been observed and the record is not censored}.
\end{cases}
\label{eq:delta}
\end{equation}

Moreover, assume that each observation is described by a set of covariates or features $\mathbf{X}_i=\{x_{i_1},\ldots,x_{i_N}\}$.
Thus, each record $i$ comprises the variables $\mathbf{X}_i$, $t_i$, and $\delta_i$ as specified in~\eqref{eq:delta}.
If $\delta_k=1$, the event is confirmed to have occurred at time $t_k$ and $y_{k_j} = 0$ for all $j \geq{k}$.
Conversely, if $\delta_k=0$, it implies that the fate of the instance beyond time $t_k$ is unknown, making $y_{k_j}$ missing for $j \geq{t_k}$.
This scenario of censored examples introduces missing data, which makes such datasets apt for semi-supervised learning methodologies."

@@

As a main part of the \gls{vspyct} model lies the oblique tree structure, which is a form of a \gls{pct} where splits are made using linear combinations of the input features rather than single features.
This allows for more complex decision boundaries, capturing intricate patterns in the data with fewer nodes compared to axis-aligned trees.
However, unlike traditional oblique trees that fix the parameters of these linear combinations, \gls{vspyct} introduces a probabilistic twist: the parameters of the splits are not single fixed values but are instead modelled as random variables with Gaussian distributions.
This probabilistic representation captures the uncertainty in the model parameters, a feature that is particularly useful in scenarios with noisy data.
The construction of the \gls{vspyct} begins with a root node and iteratively expands by adding internal nodes or leaves through a learning process based on the variational Bayes method.
Each node represents a binary decision point where instances are split into two child nodes.
At each internal node, a variational Bayes approach is used to learn the parameters of an oblique split.
This involves optimising the parameters of a linear model that best divides the data to minimise the ELBO.
The linear model is defined as:

\begin{equation}
f(\mathbf{x})=\sigma \left(\mathbf{w}^T \mathbf{x} + b \right)
\label{eq:linear_model}
\end{equation}

where $\mathbf{x} \in \mathbb{R}^N$ is the feature vector, $\mathbf{w} \in \mathbb{R}^N$ represents the weights, $b \in \mathbb{R}$ is the bias, and $\sigma(\cdot)$ denotes the sigmoid function ensuring the output lies in $(0, 1)$, indicating the probability of branching to the right child node.
The learning process involves defining a prior distribution over the weights and biases, typically Gaussian, and updating this to a posterior distribution given the observed data through variational inference.
The ELBO is optimised using stochastic gradient descent, with the learning rate $\alpha$ and potentially variable batch sizes $B$.
The split at each node is determined by evaluating the learned linear model on the descriptive data.
Instances are split based on the probability threshold of $0.5$, with those having a probability greater than $0.5$ moving to the right child node and the rest to the left.
For leaf nodes, a prototype is computed as the mean of the target variables for the instances assigned to the leaf.
This prototype represents the prediction for instances that fall into the leaf during the prediction phase.

@@

The learning of these Gaussian-distributed parameters is achieved through a Variational Bayes (VB) method.
In the context of \gls{vspyct}, the VB method is employed to learn the Gaussian distributions representing the parameters of the oblique splits.
The process involves defining a prior distribution over the parameters and then updating this prior to a posterior distribution given the observed data.
This is done by optimising the ELBO, which balances the fit of the model to the data with the complexity of the model.
The procedure for learning a split using the VB method is given in Algorithm~\ref{alg:learn_split}.

\begin{algorithm}[h!]
\caption{Variational Learning of Split Parameters in \gls{vspyct}}
\label{alg:learn_split}
\begin{algorithmic}[1]

\State \textbf{Input:} Dataset $\mathcal{D}$, features $\mathbf{X}$, targets $\mathbf{Y}$, parameters $\theta$, epochs $E$, batch size $\beta$, learning rate $\lambda$, selection probability $\sigma$
\State \textbf{Output:} Parameters $\Theta$, guide $\mathcal{G}$

\Procedure{LearnSplit}{$\mathcal{D}$, $\mathbf{X}$, $\mathbf{Y}$, $\theta$, $E$, $\lambda$, $\beta$, $\sigma$}
    \State $\Phi \sim \text{Bernoulli}(\sigma)$ \Comment{Select attributes based on probability $\sigma$}
    \State $\mathbf{X}_\Phi \leftarrow$ subset of $\mathbf{X}$ using $\Phi$
    \State Model $\mathcal{M} \leftarrow \text{Define Impurity}(\dim(\mathbf{X}_\Phi), 1, \theta)$
    \State Guide $\mathcal{G} \leftarrow \text{Define AutoDiagonalNormal}(\mathcal{M})$
    \State Define optimization parameters $\alpha \leftarrow \{\text{"lr"}: \lambda\}$
    \State Initialise optimizer $\text{Optimizer} \leftarrow \text{Adam}(\alpha)$
    \State Early stopping mechanism $\text{EarlyStopping} \leftarrow \text{Initialize}(\beta)$

    \State Initialise loss list $L$
    \State Calculate number of batches $B \leftarrow \lceil \frac{|\mathcal{D}|}{\beta} \rceil$

    \For{$e \in \{1, \dots, E\}$}
        \State Initialise training loss $\mathcal{L}_{\text{train}} \leftarrow 0$
        \For{$b \in \{1, \dots, B\}$}
            \State $(\mathbf{X}_b, \mathbf{Y}_b) \leftarrow \text{Get Batch}(\mathbf{X}_\Phi, \mathbf{Y}, b, \beta)$
            \State Update training loss $\mathcal{L}_{\text{train}} \leftarrow \mathcal{L}_{\text{train}} + \text{SVI Step}(\mathcal{M}, \mathcal{G}, \mathbf{X}_b, \mathbf{Y}_b)$
        \EndFor
        \State Append $\mathcal{L}_{\text{train}}$ to $L$
        \If{$\text{EarlyStopping}(\mathcal{L}_{\text{train}})$}
            \State \textbf{break}
        \EndIf
    \EndFor
    \State \Return $\mathcal{M}$, $\mathcal{G}$
\EndProcedure

\begin{algorithm}[h!]
\caption{Impurity}
\begin{algorithmic}[1]

\Procedure{Impurity}{$\xi$, $\zeta$, $\theta$}
    \State $\mathbf{w}, b \sim \mathcal{N}(0, I)$ \Comment{Weight and bias priors}
    \State $\rho(\mathbf{x}) \gets \sigma(\mathbf{w}^\top \mathbf{x} + b)$ \Comment{Right selection probability}
    \State $\lambda(\mathbf{x}) \gets 1 - \rho(\mathbf{x})$ \Comment{Left selection probability}
    \State Compute $\text{Var}_\rho(\mathbf{Y})$ and $\text{Var}_\lambda(\mathbf{Y})$ \Comment{Weighted variances}
    \State $\Omega \gets \lambda \cdot \text{Var}_\lambda(\mathbf{Y}) + \rho \cdot \text{Var}_\rho(\mathbf{Y}) + \|\mathbf{w}\|_{0.5}$
%    \State Observe $\Omega$ with likelihood $\mathcal{N}(0, 1)$
    \State \textbf{return} $\Omega$
\EndProcedure

@@

The procedure for making a prediction for a new instance using the \gls{vspyct} model is given in Algorithm~\ref{alg:make_pred}.
It involves traversing the tree from the root to a leaf, at each internal node using the variational posterior to sample parameters and compute the probability of branching right or left.
The prediction of the reached leaf's prototype is returned as the output.
Operationally, when a new data point is to be classified or a prediction is to be made, it traverses the tree from the root to a leaf.
At each node, the oblique split is determined by evaluating the linear combination of the input features, with parameters sampled from the learned Gaussian distributions.
This probabilistic approach to decision-making within the tree allows for a more nuanced interpretation of the data, accounting for uncertainty in the model parameters.

\begin{algorithm}
\caption{\gls{vspyct} Prediction Process}
\label{alg:make_pred}
\begin{algorithmic}[1]

\State \textbf{Input:} Feature vector $\mathbf{x}$, decision tree $T$ rooted at node $\Omega$
\State \textbf{Output:} Prediction $\hat{y}$

\Function{Predict}{$\Omega$, $\mathbf{x}$}
    \If{$\Omega$ is a leaf}
        \State \Return $\Omega.prototype$
    \Else
        \State $\Theta \gets$ \Call{SampleParameters}{$\Omega.split\_model$, $\Omega.guide$}
        \State Define sigmoid function $\rho(\Theta, \mathbf{x}) = \frac{1}{1 + \exp(-\Theta^\top \mathbf{x})}$
        \If{$\rho(\Theta, \mathbf{x}) \leq 0.5$}
            \State \Return \Call{Predict}{$\Omega.left$, $\mathbf{x}$}
        \Else
            \State \Return \Call{Predict}{$\Omega.right$, $\mathbf{x}$}
        \EndIf
    \EndIf
\EndFunction

>>>>

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{pythontex}
\usepackage{graphicx}

\usepackage[acronym]{glossaries}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{bs}{BS}{Brier Score}
\newacronym{pes}{PES}{Public Employment Services}
\newacronym{pex}{PEX}{probability of exit}
\newacronym{spyct}{SPYCT}{oblique predictive clustering tree}
\newacronym{vspyct}{VSPYCT}{variational oblique predictive clustering tree}
\newacronym{opct}{OPCT}{option predictive clustering tree}
\newacronym{sslmtr}{SSL-MTR}{semi-supervised multi-target regression}
\newacronym{pct}{PCT}{predictive clustering tree}
\newacronym{sop}{SOP}{structured output prediction}
\newacronym{fp}{FP}{False Positive}
\newacronym{fn}{FN}{False Negative}
\newacronym{tpr}{TPR}{True Positive Rate}
\newacronym{fpr}{FPR}{False Positive Rate}
\newacronym{auc}{AUC}{Area under the ROC curve}
\newacronym{tp}{TP}{True Positives}
\newacronym{tn}{TN}{True Negatives}
\newacronym{ecdf}{ECDF}{Empirical Cumulative Distribution Function}
\newacronym{isce}{ISCED}{International Standard Classification of Education}
\newacronym{kl}{KL}{Kullback–Leibler}
\newacronym{elbo}{ELBO}{Evidence Lower Bound}

@@

\begin{document}
\begin{frontmatter}
	\title{Variational oblique predictive clustering trees}
	\author[1,2]{Viktor Andonovikj\corref{cor1}}
	\ead{viktor.andonovikj@ijs.si}
	\author[1]{Sašo Džeroski}
	\ead{saso.dzeroski@ijs.si}
	\author[1,3]{Biljana Mileva Boshkoska}
	\ead{biljana.mileva@ijs.si}
	\author[1]{Pavle Boškoski}
	\ead{pavle.boskoski@ijs.si}
	\address[1]{Jožef Stefan Institute, Jamova cesta 39, 1000 Ljubljana, Slovenia}
	\address[2]{Jožef Stefan International Postgraduate School, Jamova cesta 39, 1000 Ljubljana, Slovenia}
	\address[3]{Faculty of Information Studies in Novo mesto, Ljubljanska cesta 31b, 8000 Novo mesto, Slovenia}
	\cortext[cor1]{Corresponding author.} 
\begin{abstract}

\Glspl{spyct} are semi-supervised multi-target prediction models mainly used for \gls{sop} problems.
They are computationally efficient and when combined in ensembles they achieve state-of-the-art results.
However, one major issue is that it's challenging to interpret an ensemble of \glspl{spyct} without the use of a model-agnostic method.
We propose variational oblique predictive clustering trees, which address this challenge.
The parameters of each split node are treated as random variables, described with a probability distribution, and they are learned through the Variational Bayes method.
We evaluate the model on several benchmark datasets of different sizes.
The experimental analyses show that a single \gls{vspyct} achieves competitive, and sometimes better predictive performance than the ensemble of standard \glspl{spyct}.
We also present a method for extracting feature importance scores from the model.
Finally, we present a method to visually interpret the model's decision making process through analysis of the relative feature importance in each split node.
\end{abstract}
%\maketitle
\end{frontmatter}
\glsresetall

@@

Our proposed \gls{vspyct} framework diverges from traditional ensemble approaches by directly incorporating Bayesian principles into oblique decision trees' architecture.
This model avoids conventional ensemble strategies, which aggregate multiple trees' outputs, for a probabilistic treatment of model parameters, offering refined mechanisms for uncertainty quantification and predictive performance.
The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, modelling these parameters as random variables with probability distributions.
This represents a paradigm shift toward a probabilistic understanding of decision-making processes within tree-based models, significantly enhancing interpretability, especially in uncertainty estimation domains.
It sets a new benchmark for interpretable, efficient, and reliable machine learning models.

@@

Our proposed \gls{vspyct} model addresses these critical challenges by integrating Bayesian principles directly into the predictive clustering tree framework.
%Unlike traditional ensemble strategies, \glspl{vspyct} employs a probabilistic treatment of model parameters through the variational Bayes method, which enhances interpretability and provides inherent uncertainty quantification.
This approach maintains the clarity and simplicity of a single tree model while achieving state-of-the-art performance, thus offering a significant improvement over existing methods.
By introducing uncertainty quantification directly into the decision-making process, \gls{vspyct} enhances the reliability and applicability of the model across various domains where decision certainty is crucial.
%This advancement bridges the gap between theoretical developments and practical applications, marking a significant step forward in making complex decision trees more accessible, interpretable, and effective.

\section{The variational Bayes method}

@@

In conducting a stochastic analysis, the process of inferring model parameters hinges on applying Bayes' theorem.
For observations $x$ produced by a system defined by parameters $\theta$, Bayes' theorem is given as:

@@

This integral rarely yields a closed-form solution.
For cases with multiple dimensions, the computational burden often renders Monte Carlo methods infeasible.
The VB approach offers an approximation to this issue by closely mimicking the true posterior $p(\theta|x)$ with an approximate distribution $q_{\omega^*}(\theta)$. This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterized by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
Typically, the mean-field variational family is utilised, assuming independence among the parameters $\omega$~\citep{Blei_2017}. 
To find the optimal $\omega^*$, one minimises the \gls{kl} divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:

@@

Given the unknown nature of the true posterior, a rearrangement is needed to compute the \gls{kl} divergence, simplifying to~\citep[Chapter~10]{murphy2023probabilistic}:

\begin{align*}
KL(q_\omega(\theta) \parallel p(\theta|x)) = - & \mathbb{E}_q\left[\log p(x,\theta) - \log q_\omega(\theta)\right] + \log p(x)\\
& \underbrace{\phantom{\mathbb{E}_q\left[\log p(x,\theta) - \log q_\omega(\theta)\right]}}_{\text{ELBO}}
\label{eq:kl_divergence}
\end{align*}

Maximising the \gls{elbo} serves to inversely minimise the \gls{kl} divergence between the approximate posterior \(q_\omega(\theta)\) and the true posterior \(p_\omega(\theta)\).
In the decomposition of the KL divergence, the term \(\log p(x)\), known as the marginal likelihood or evidence, remains constant with respect to the variational parameters \(\theta\).
Since this term does not involve \(\theta\), it does not influence the optimisation of the variational distribution and is thus excluded from the \gls{elbo} maximisation process.
By maximising the \gls{elbo}, we aim to tighten the bound provided by the \gls{kl} divergence, effectively making the variational posterior a better approximation of the true posterior.
Typically, the \gls{elbo} maximisation criterion is non-convex, with no assurance of global extremum convergence by optimisation algorithms~\citep[Chapter~10]{murphy2023probabilistic}.
Various algorithms have been proposed for this challenge, including variational EM \citep{bernardo2003variational}, stochastic variational inference~\citep{hoffman2013stochastic, sashank2018convergence}, amortised variational inference~\citep{gershman2014amortized, le2017inference}, and semi-amortised inference~\citep{kim2018semi}.

The adoption of an approximate variational distribution introduces bias contingent on the chosen variational family $\mathcal{Q}$, necessitating a decision grounded in empirical evidence or expert knowledge.
This bias arises because the family \(\mathcal{Q}\) may not be capable of capturing the true complexity of the posterior distribution \(p_\omega(\theta)\).
As a result, the accuracy of the approximation depends significantly on how well \(\mathcal{Q}\) aligns with the true underlying distribution.
The choice of \(\mathcal{Q}\) thus not only affects the efficiency of the inference but also the quality and reliability of the model.
Despite this bias, VB's computational efficiency significantly surpasses traditional methods like Markov Chain Monte Carlo (MCMC).
In our case, the ADAM optimiser~\citep{kingma2014adam} is used to facilitate the \gls{elbo} optimisation.

@@

As the foundation of the \gls{vspyct} model, the underlying architecture closely follows that of the \gls{spyct} model, particularly in its tree construction process.
Similar to \gls{spyct}, \gls{vspyct} constructs a decision tree with an oblique structure, where each split is defined by a linear combination of input features, rather than relying on single-feature thresholds as in axis-aligned trees.

The key distinction between \gls{vspyct} and \gls{spyct} lies in the optimisation step used to determine the parameters of these oblique splits.
While \gls{spyct} utilises a deterministic approach to optimise split parameters, \gls{vspyct} introduces a probabilistic framework through variational Bayes (VB) optimisation.
In \gls{vspyct}, the parameters of the linear combination at each split—specifically, the weight vector \(\mathbf{w}\) and the bias term \(b\)—are modelled as random variables with Gaussian distributions.
This probabilistic representation of split parameters not only enhances the model's ability to handle uncertainty and noise in the data but also provides a measure of predictive uncertainty.

The construction of the \gls{vspyct} begins with a root node and proceeds iteratively, expanding the tree by adding internal nodes or leaf nodes in a manner identical to \gls{spyct}. Each node in the tree represents a binary decision point, where instances are split based on the oblique hyperplane defined by the linear model:

\begin{equation}
f(\mathbf{x}) = \sigma \left(\mathbf{w}^\top \mathbf{x} + b \right)
\label{eq:linear_model}
\end{equation}
where \(\mathbf{x} \in \mathbb{R}^N\) is the feature vector, \(\mathbf{w} \in \mathbb{R}^N\) represents the weight vector, \(b \in \mathbb{R}\) is the bias term, and \(\sigma(\cdot)\) denotes the sigmoid function, which ensures the output is a probability, determining whether an instance is sent to the right or left child node.

While the tree structure and the decision-making process follow the same principles as in \gls{spyct}, the optimisation of \(\mathbf{w}\) and \(b\) in \gls{vspyct} differs fundamentally.
In \gls{vspyct}, these parameters are not fixed but are instead inferred through variational Bayes optimisation.
The introduction of this probabilistic element distinguishes \gls{vspyct} from \gls{spyct}, enabling it to maintain the interpretability of a single decision tree while incorporating the robustness typically associated with ensemble methods.

\begin{comment}
\pbnote{Why noisy data, what if we have perfect data but wrong model}
\vanote{If we have perfect data - then even a deterministic model will do well for the prediction. We choose the type of model before we do the optimisation. If we choose a wrong model - e.g. a decision tree where it is simply a mismatch for the type of problem at hand - we will have bad predictive performance, regardless of the modelling framework (deterministic or probabilistic). But if we choose the right type of model, and face noisy data at test time - then a probabilistic model would capture the predictive uncertainty in its parameter distributions.}
\end{comment}

@@

The goal is to update these prior distributions to approximate the posterior distributions given the observed data, thereby capturing the uncertainty in the model parameters.
The weights \(\mathbf{w}\) and bias \(b\) in the linear model, as in~\eqref{eq:linear_model}, are initially modelled with Gaussian prior distributions:

\begin{equation}
\mathbf{w} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}), \quad b \sim \mathcal{N}(0, 1)
\end{equation}

The VB method aims to approximate the true posterior distributions of these parameters with a variational distribution \(q(\mathbf{w}, b|\mathcal{D})\).
We assume a simpler distributional form - a diagonal Gaussian:

\begin{equation}
q(\mathbf{w}, b|\mathcal{D}) = \mathcal{N}(\mathbf{w}|\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w) \mathcal{N}(b|\mu_b, \sigma_b^2)
\end{equation}

where \(\boldsymbol{\mu}_w\), \(\boldsymbol{\Sigma}_w\), \(\mu_b\), and \(\sigma_b^2\) are the parameters of the variational distribution, learned through the optimisation process.

The optimisation of the variational parameters is performed by maximising the \gls{elbo}, which serves as a lower bound to the log marginal likelihood \(\log p(\mathcal{D})\).
The \gls{elbo} is given by:

\begin{equation}
\mathcal{L}(\boldsymbol{\mu}, \boldsymbol{\Sigma}) = \mathbb{E}_{q(\mathbf{w}, b|\mathcal{D})}\left[\log p(\mathcal{D}|\mathbf{w}, b)\right] - KL\left(q(\mathbf{w}, b|\mathcal{D}) \parallel p(\mathbf{w}, b)\right)
\label{eq:elbo}
\end{equation}

The first term represents the expected log-likelihood of the data under the variational distribution, while the second term is the Kullback-Leibler (KL) divergence between the variational distribution and the prior. The \gls{elbo} is maximised using the ADAM optimiser, with the goal of refining the variational parameters \(\boldsymbol{\mu}\) and \(\boldsymbol{\Sigma}\) to closely approximate the true posterior.

The quality of a split is assessed using an impurity function, which is defined as:

\begin{equation}
\text{Impurity} = \sum_{i=1}^{N} \left(\rho_i \cdot \text{Var}_{\rho}(\mathbf{Y}) + \lambda_i \cdot \text{Var}_{\lambda}(\mathbf{Y})\right)
\end{equation}

where \(\rho_i = \sigma(\mathbf{w}^\top \mathbf{x}_i + b)\) is the probability of an instance being routed to the right child node, and \(\lambda_i = 1 - \rho_i\) is the probability of routing to the left child node.
The variances \(\text{Var}_{\rho}(\mathbf{Y})\) and \(\text{Var}_{\lambda}(\mathbf{Y})\) represent the variance of the target variables in the right and left child nodes, respectively, weighted by their selection probabilities.

The objective is to minimise the impurity by optimising the parameters \(\mathbf{w}\) and \(b\) using the VB method, thereby ensuring that the split at each node effectively separates the data into homogeneous subsets.

The overall procedure for learning a split in \gls{vspyct} is detailed in Algorithm~\ref{alg:learn_split_vb}.
The algorithm involves initialising the variational parameters, performing Monte Carlo sampling to estimate the \gls{elbo}, and iteratively updating the parameters until convergence.

\begin{algorithm}[h!]
\caption{Variational Learning of Split Parameters in \gls{vspyct} with Impurity Minimization}
\label{alg:learn_split_vb}
\begin{algorithmic}[1]
    \State \textbf{Input:} $\mathcal{D} = \{\mathbf{X}, \mathbf{Y}\}$ (dataset), $\theta = \{\mathbf{w}_0, b_0\}$ (initial parameters), $E$ (epochs), $\lambda$ (learning rate), $\beta$ (batch size), $\sigma$ (selection probability)
    \State \textbf{Output:} $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ (variational parameters)

    \Procedure{LearnSplit}{$\mathcal{D}, \theta, E, \lambda, \beta, \sigma$}
        \State Initialize $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ \Comment{Initialize variational parameters}
        \State Initialize optimizer: $\alpha \gets \lambda$
        \For{$e \in \{1, \dots, E\}$} \Comment{Iterate over epochs}
            \For{each mini-batch $\mathcal{B} \subseteq \mathcal{D}$ of size $\beta$}
                \State Sample $\mathbf{w} \sim \mathcal{N}(\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w)$, $b \sim \mathcal{N}(\mu_b, \sigma_b^2)$
                \State Compute impurity $\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)$ as:
                \[
                \Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b) = \sum_{\text{right}} \rho(\mathbf{w}^\top \mathbf{x} + b) \cdot \text{Var}_\rho(\mathbf{Y}) + \sum_{\text{left}} (1 - \rho(\mathbf{w}^\top \mathbf{x} + b)) \cdot \text{Var}_\lambda(\mathbf{Y})
                \]
                \State Set the observed impurity value: $\Omega_\text{obs} = \frac{\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)}{2}$
                \State Condition on the observed impurity using:
                \[
                \log p(\Omega_\text{obs}|\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b))
                \]
                \State Compute ELBO $\mathcal{L}(\mathcal{B}; \Theta)$ as per~\eqref{eq:elbo}
                \State Compute gradient $\nabla_{\Theta} \text{ELBO}$
                \State Update $\Theta \gets \Theta + \alpha \nabla_{\Theta} \text{ELBO}$ \Comment{Gradient descent update}
            \EndFor
            \If{early stopping criterion satisfied}
                \State \textbf{break}
            \EndIf
        \EndFor
        \State \textbf{return} $\Theta$
    \EndProcedure
\end{algorithmic}
\end{algorithm}

@@

The procedure for making a prediction for a new instance using the \gls{vspyct} model is detailed in Algorithm~\ref{alg:make_pred}.
The process involves traversing the tree from the root node to a leaf node, with decisions at each internal node being made probabilistically based on the parameters sampled from the learned variational posterior distributions.
The final prediction is determined by the prototype value stored in the reached leaf node.

Mathematically, at each internal node, the oblique split is determined by evaluating the linear combination of input features, with the parameters \(\Theta = (\mathbf{w}, b)\) drawn from their respective variational posterior distributions. The decision to branch left or right is based on the probability computed using the sigmoid function applied to this linear combination. This method accounts for the uncertainty in the model parameters, providing a probabilistic framework for decision-making.

The prediction \(\hat{y}\) returned by the model is the prototype of the reached leaf node, where the prototype is defined as the mean of the target values associated with the training instances that were routed to that leaf during training.

\begin{equation}
\hat{y} = \frac{1}{|\mathbf{Y}|} \sum_{i=1}^{|\mathbf{Y}|} y_i
\end{equation}
where \( y_i \) represents the target value of the \(i\)-th instance in \(\mathbf{Y}\).
This value $\hat{y}$ serves as the prediction for any new instance that reaches the particular leaf node.

\begin{algorithm}[h!]
\caption{Prediction Process in \gls{vspyct} using Monte Carlo Sampling}
\label{alg:make_pred}
\begin{algorithmic}[1]

\State \textbf{Input:} Feature vector $\mathbf{x}$, decision tree $T$ rooted at node $\Omega$, number of samples $M$
\State \textbf{Output:} Prediction $\hat{y}$

\Function{MC\_Predict}{$\Omega$, $\mathbf{x}$, $M$}
    \State Initialize $\hat{y}_{\text{sum}} = 0$
    \For{$m = 1$ to $M$}
        \State $\hat{y}_{\text{sum}} \gets \hat{y}_{\text{sum}} + \text{Predict}(\Omega, \mathbf{x}, m)$
    \EndFor
    \State \Return $\hat{y} = \frac{\hat{y}_{\text{sum}}}{M}$
\EndFunction
